Understanding Complex Patterns in Social, Geographic, and Economic Inequities in COVID-19 Mortality at the County Level in the US Using Generalized Additive Models

### 2022-11-27 manuscript/manuscript.rmd
updating; first draft done of discussion + conclusions

  \noindent Abbreviations used: United States (US), Coronavirus Disease 2019 (COVID-19), Generalized
  Additive Model (GAM), Centers for Disease Control and Prevention (CDC)
bibliography: bibliography.bib
csl: apa.csl
link-citations: true
urlcolor: blue
linkcolor: black
header-includes: 
  - \usepackage{pdflscape}
  - \newcommand{\blandscape}{\begin{landscape}}
  - \newcommand{\elandscape}{\end{landscape}}
---

@@

Prior literature has demonstrated that geography has played a significant role
in the spread and impact of of COVID-19. Methods employed to highlight the role
geography plays have included quantile regression [@sigler_socio-spatial_2021],
Besag-York-Molli\'e mixed models [@whittle_ecological_2020], spatial cluster analysis
[@sugg_mapping_2021], geographically weighted regression [@mollalo_gis-based_2020; @park_covid-19_2021], and others.

@@

  * Counts of COVID-19 deaths [@the_new_york_times_coronavirus_2021].
  * Population size estimates for 2020 from the U.S. Census (Redistricting 
  File Public-Law 94-171 Dataset) [@us_census_bureau_2020_2021].
  * Median age, median household income, racial/ethnic composition, population
  density, percent below the federal poverty line, and number of households with
  high (\$100k+)/low (<\$25k) household income by racial/ethnic group from the
  2014-2019 5-year American Community Survey [@us_census_bureau_american_2020]
  through the `tidyverse` R package [@walker_tidycensus_2022]. 
  * Votes cast in the 2020 presidential election [@mit_election_data_and_science_lab_county_2022]

@@

Traditional generalized linear models fit a regression model using the
functional relation $g(\mu_i) = \mathbf X_i \beta$ where Examples of common link
functions include the identity function for linear regression, $\log$ for
Poisson regression, and the $\text{logit}$ function for logistic regression.

In contrast, generalized additive models (GAMs) improve upon generalized linear models by
allowing for the fitting of smooth functions that transform the $x$ variables.
This is a convenient means to include nonlinear relationships between the
outcome and predictor variables. Whereas a generalized linear model may have looked like 
$$ g(\mu_i) = \beta_0 + \beta_1 x_{i1} + \beta_2 x_{i2} + \beta_3 x_{i3} ...$$ 
a generalized additive model could look like 
$$ g(\mu_i) = \mathbf A_i \mathbf \theta + f_1(x_{i1}) + f_2(x_{i2}) + f_3(x_{i3}, x_{i4}) + \dots$$
where $\mu_i \equiv \mathbb E(Y_i)$, $\mathbf A_i$ is a row of the model matrix 
for any strictly non-parametric model components, $\mathbf \theta$ is the corresponding
parameter vector, and the $f_j$ are smooth functions of the covariates $x_k$ [@wood_generalized_2017].

@@

A negative binomial model in the context of modeling count data can be written
$$y_i \sim \text{NegativeBinomial}(u_i \theta_i, \phi),$$
where $y_i$ are the counts observed, $u_i$ is the exposure, $\theta_i$ are
the rates, and $\phi$ determines the amount of overdispersion. The rates
are modeled as $\theta_i = e^{X_i \beta}$ where $X_i$ are the observed
covariates and $\beta$ are the coefficients on the covariates corresponding
to log rate ratios. The logarithm of the exposure, $\log(u_i)$ is often called 
(and later herein referenced as) the offset. In epidemiological contexts, the offset 
is often representative of the amount of person-time during which observations were recorded. Whereas the Poisson model holds that $\text{var}(y) = \mu$ where $\mu$
is the average rate, the negative binomial model instead assumes that
$\text{var}(y) = \mu + \mu^2/\phi$. The Poisson model
is a special case of the negative binomial model when $\phi \to \infty$.

@@

Besides age, which is the greatest risk factor for COVID-19 mortality, we have
included the following variables as covariates of interest:

@@

The GAM models are fit with the following structure using 
the `gam` function from the `mgcv` package [@wood_mgcv_2022]:

@@

The above structure is used to estimate models for our different `covariate`
variables of interest: median income, percent in poverty, the ICE for racialized
economic segregation, political lean.  Whereas the other models treat median age
as a variable to adjust for, the model presenting the effect of median age over
time treats median age as the main covariate of interest including it as the 
`te(covariate, date, d=c(1,1))` and dropping the `s(median_age)` term which is 
redundant. Results of these models are summarized in \@ref(fig:one-variable-over-time)
where the additional COVID-19 mortality associated with each covariate is visualized.

@@

Another non-spatial application of GAMs to describe the distribution of 
COVID-19 mortality in the US over time is to consider the associations of 
mortality with three-way interactions of time and two covariates taken together. 
In the following example, the formula used is `deaths ~ te(date, ICEraceinc, median_age, d=c(1,1,1))`.  Again, the `d` argument specifying the marginal 
basis dimensions is used indicate that each of the `date`, `ICEraceinc`, and 
`median_age` measures are in different units and should not be smoothed assuming
that a one unit difference in one variable is comparable to a one unit difference
in another variable.

```{r two-variables-over-time, fig.cap="The changing interacted effect of the ICE for racialized economic segregation and age over time. A) May 2020, B) November 2020, C) May 2021, D) October 2021, E) February 2022, F) August 2022. An animated version is available online at [https://github.com/ctesta01/covid.gradient.estimation/blob/main/analysis/05_two_variables_at_a_time/animation_ICEraceinc_age/readme.md](https://github.com/ctesta01/covid.gradient.estimation/blob/main/analysis/05_two_variables_at_a_time/animation_ICEraceinc_age/readme.md)"}
knitr::include_graphics(here("analysis/05_two_variables_at_a_time/animation_ICEraceinc_age/panel_figure.png"))
```

## Application 2: Spatiotemporal Smoothing

By fitting GAMs with a tensor-product smoothing term on latitude, longitude, and
time we can estimate a spatiotemporally smoothed trend in COVID-19 mortality. To
do this, the GAM is constructed similarly as in Application 1 but with the
formula specified as `deaths ~ te(latitude, longitude, time, d=c(2,1))` where
`d=c(2,1)` indicates that `latitude` and `longitude` share the same dimensions
(i.e., both are spatial and in units of degrees) while the `time` data are in separate
units.

Using GAMs to present spatiotemporally smoothed estimates of COVID-19 mortality
allows for the synthesis of trends over time. Whereas the raw rates of COVID-19
mortality are noisy due to 72% of the land area in the US being classified as
rural and low population density, the spatiotemporally smoothed GAM estimates
use population weighting via the offset specified to estimate a latent surface
that represents localized mortality rate averages in spatial coordinates and in
time.

@@

```{r spatiotemporal-panel-figure, fig.cap = "Spatiotemporally smoothed COVID-19 mortality estimates from a GAM fit to data from March 2020-August 2022. Panels show months selected to highlight the changing spatial patterns of COVID-19 mortality risk over time. An animated version is available online at []()"}
knitr::include_graphics(here("analysis/09_spatiotemporal_models/temporal_panel_figure.png"))
```

## Application 3: Estimating Covariate Effects Adjusted for Spatiotemporal Autocorrelation

### Evaluating Spatial and Temporal Autocorrelation

```{r figure county neighboring graph, fig.cap = "A) The spatial neighbor graph for counties in the US. In this graph, county centroids are connected if the two counties are adjacent to one another (i.e., share a boundary). This and similar graphs that represent neighbor relationships or geographic proximity are often used in modeling contexts to create spatial weights that account for which regions help to explain a given region's observed data.", fig.show='hold', out.width = '50%', fig.env = "figure*"} 
desired_caption <- 
  " B) Moran's scatter plot \\citep{anselin_simple_1996} depicts for each region (in our case, counties) on the x-axis the region rate and on the y-axis the average of the region's neighbors' rates. This diagnostic plot provides a simple means to assess to what degree neighboring regions share similar rates. In a setting where regions have high correlation, Moran's scatter plot will exhibit a high degree of positive correlation."
knitr::include_graphics(here("analysis/08_spatiotemporal_autocorrelation/county_neighboring_relations.png")) 
knitr::include_graphics(here("analysis/08_spatiotemporal_autocorrelation/morans_plot.png"))
```

```{r figure 4, fig.cap = ""}
knitr::include_graphics(here("analysis/08_spatiotemporal_autocorrelation/temporal_acf.png"))
```

## Regression Modeling Results

We used GAMs to describe the following:

  * the changing associations between individual area-based measures
  and COVID-19 mortality over time;
  * the changing associations between bivariate area-based measures
  and COVID-19 mortality over time; 
  * associations with area-based measures and COVID-19 mortality after
  controlling for spatio-temporal autocorrelation.

Additionally, as sensitivity analyses, random forest models were fit to compare
with the coefficient estimates from the GAMs.

```{r figure 6, fig.cap = ""}
knitr::include_graphics(here("analysis/09_spatiotemporal_models/panel_figure.png"))
```

# Discussion

## Limitations

  * Ecological fallacy 
  * Age effects
  * Fixed population denominators -> real-time denominator estimates
  * Vaccination and mobility effects
  * Equal Smoothing Across All Areas
  * Soap Film Smoothers

@@

# References

<!-- 1 million covid deaths: 
https://coronavirus.jhu.edu/from-our-experts/u-s-officially-surpasses-1-million-covid-19-deaths -->

<!-- nearly 225,000 Americans have died since the start of 2022
https://abcnews.go.com/Health/225000-americans-died-covid-19-start-2022-viral/story?id=90339579 -->

<!-- 
vaccine timeline
https://www.mayoclinic.org/coronavirus-covid-19/history-disease-outbreaks-vaccine-timeline/covid-19
-->

<!--
less than 4% of americans had gotten the booster 1 month into availability
https://www.nbcnews.com/health/health-news/updated-covid-booster-shots-doses-administered-cdc-rcna48960
-->

<!-- 2020 ACS data are experimental 
https://www.census.gov/newsroom/press-releases/2021/experimental-2020-acs-1-year-data.html -->

<!--
Spatial-temporal generalized additive model for modeling COVID-19 mortality risk in Toronto, Canada
https://www.ncbi.nlm.nih.gov/pmc/articles/PMC8257405/ -->

<!-- 
Using GAM functions and Markov-Switching models in an evaluation framework to assess countries’ performance in controlling the COVID-19 pandemic
https://bmcpublichealth.biomedcentral.com/articles/10.1186/s12889-021-11891-6 -->

<!--
Statistical modelling of COVID-19 data: Putting generalized additive models to work
https://journals.sagepub.com/doi/full/10.1177/1471082X221124628
-->

<!--
Inference following multiple imputation for generalized additive models: an investigation of the median p-value rule with applications to the Pulmonary Hypertension Association Registry and Colorado COVID-19 hospitalization data
https://bmcmedresmethodol.biomedcentral.com/articles/10.1186/s12874-022-01613-w
-->

<!--
Generalized Additive Models to Capture the Death Rates in Canada COVID-19
https://link.springer.com/chapter/10.1007/978-3-030-85053-1_7 -->

<!--
Contribution of Two-Dose Vaccination Toward the Reduction of COVID-19 Cases, ICU Hospitalizations and Deaths in Chile Assessed Through Explanatory Generalized Additive Models for Location, Scale, and Shape
https://www.frontiersin.org/articles/10.3389/fpubh.2022.815036/full -->

<!-- 
Combining graph neural networks and spatio-temporal disease models to improve the prediction of weekly COVID-19 cases in Germany
https://www.nature.com/articles/s41598-022-07757-5 -->

<!-- 
Association between mobility patterns and COVID-19 transmission in the USA: a mathematical modelling study
https://www.sciencedirect.com/science/article/pii/S1473309920305533 -->

<!-- 
Examining spatiotemporal evolution of racial/ethnic disparities in human mobility and COVID-19 health outcomes: Evidence from the contiguous United States
https://www.sciencedirect.com/science/article/pii/S2210670721007721 -->

<!--
Social, economic, and environmental factors influencing the basic reproduction number of COVID-19 across countries
https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0252373 -->

<!-- 
https://asbates.rbind.io/2019/05/03/gams-for-time-series/
-->

<!--
https://environmentalcomputing.net/statistics/gams/
-->

<!--
https://petolau.github.io/Analyzing-double-seasonal-time-series-with-GAM-in-R/
-->

>>>>

  \noindent Abbreviations used: United States (US), Coronavirus Disease 2019 (COVID-19), Generalized
  Additive Model (GAM), Centers for Disease Control and Prevention (CDC), Index of Concentration at the Extremes (ICE)
bibliography: bibliography.bib
csl: apa.csl
link-citations: true
urlcolor: blue
linkcolor: black
header-includes: 
  - \usepackage{pdflscape}
  - \newcommand{\blandscape}{\begin{landscape}}
  - \newcommand{\elandscape}{\end{landscape}}
---

@@

Prior literature has demonstrated that the spread and impact of COVID-19 has
varied geography over time reflecting dynamics in the timing of introduction and
transmission events. Methods employed to highlight the geographic patterning in
COVID-19 outcomes have included quantile regression
[@sigler_socio-spatial_2021], Besag-York-Molli\'e mixed models
[@whittle_ecological_2020], spatial cluster analysis [@sugg_mapping_2021],
geographically weighted regression [@mollalo_gis-based_2020;
@park_covid-19_2021], and others.

@@

  * Counts of COVID-19 deaths [@the_new_york_times_coronavirus_2021].
  * Population size estimates for 2020 from the U.S. Census [@us_census_bureau_2020_2021].
  * Median age, median household income, racial/ethnic composition, population
  density, percent below the federal poverty line, and number of households with
  high (\$100k+)/low (<\$25k) household income by racial/ethnic group from the
  2014-2019 5-year American Community Survey [@us_census_bureau_american_2020]
  through the `tidyverse` R package [@walker_tidycensus_2022]. 
  * Votes cast in the 2020 presidential election [@mit_election_data_and_science_lab_county_2022]

@@

Generalized additive models (GAMs) improve upon generalized linear models by
allowing for the fitting of smooth functions that transform the $x$ variables.
This is a convenient means to include nonlinear relationships between the
outcome and predictor variables. Whereas a generalized linear model may look like 
$$ g(\mu_i) = \beta_0 + \beta_1 x_{i1} + \beta_2 x_{i2} + \beta_3 x_{i3} ...$$ 
a generalized additive model could look like 
$$ g(\mu_i) = \mathbf A_i \mathbf \theta + f_1(x_{i1}) + f_2(x_{i2}) + f_3(x_{i3}, x_{i4}) + \dots$$
where $\mu_i \equiv \mathbb E(Y_i)$, $\mathbf A_i$ is a row of the model matrix 
for any strictly non-parametric model components, $\mathbf \theta$ is the corresponding
parameter vector, and the $f_j$ are smooth functions of the covariates $x_k$ [@wood_generalized_2017].

@@

A negative binomial model in the context of modeling count data can be written
$$y_i \sim \text{NegativeBinomial}(u_i \theta_i, \phi),$$
where $y_i$ are the counts observed, $u_i$ is the "exposure", $\theta_i$ are
the rates, and $\phi$ determines the amount of overdispersion. The rates
are modeled as $\theta_i = e^{X_i \beta}$ where $X_i$ are the observed
covariates and $\beta$ are the coefficients on the covariates corresponding
to log rate ratios. The logarithm of the exposure, $\log(u_i)$ is often called 
(and later herein referenced as) the offset. In epidemiological contexts, the offset 
is often representative of the amount of person-time during which observations were recorded. Whereas the Poisson model holds that $\text{var}(y) = \mu$ where $\mu$
is the average rate, the negative binomial model instead assumes that
$\text{var}(y) = \mu + \mu^2/\phi$. The Poisson model
is a special case of the negative binomial model when $\phi \to \infty$.

@@

Besides age, which is a strong determinant of COVID-19 mortality risk, we have
included the following variables as covariates of interest:

@@

The GAM models shown in Figure \@ref(fig:one-variable-over-time) are fit with the following structure using 
the `gam` function from the `mgcv` package [@wood_mgcv_2022]:

@@

The above structure is used to estimate models for our different `covariate`
variables of interest: median income, percent in poverty, the ICE for racialized
economic segregation, political lean.  The model presenting median age treats
median age as the main covariate of interest including it as the `te(covariate,
date, d=c(1,1))`. Results of these models are summarized in Figure
\@ref(fig:one-variable-over-time) where the additional COVID-19 mortality
associated with each covariate is visualized.

@@

A more complex non-spatial application of GAMs to describe the distribution of 
COVID-19 mortality in the US over time is to consider the associations of 
mortality with three-way interactions of time and two covariates taken together. 
In the following example, the formula used is `deaths ~ te(date, ICEraceinc, median_age, d=c(1,1,1))`.  Again, the `d` argument specifying the marginal 
basis dimensions is used indicate that each of the `date`, `ICEraceinc`, and 
`median_age` measures are in different units and should not be smoothed assuming
that a one unit difference in one variable is comparable to a one unit difference
in another variable. This approach is represented in Figure \@ref(fig:two-variables-over-time).

```{r two-variables-over-time, fig.cap="The changing interacted effect of the ICE for racialized economic segregation and age over time. A) May 2020, B) November 2020, C) May 2021, D) October 2021, E) February 2022, F) August 2022. An animated version is available online at \\url{https://github.com/ctesta01/covid.gradient.estimation/blob/main/analysis/05_two_variables_at_a_time/animation_ICEraceinc_age/readme.md}"}
knitr::include_graphics(here("analysis/05_two_variables_at_a_time/animation_ICEraceinc_age/panel_figure.png"))
```

\newpage
## Application 2: Spatiotemporal Smoothing

By fitting GAMs with a tensor-product smoothing term on latitude, longitude, and
time we can estimate a spatiotemporally smoothed trend in COVID-19 mortality. To
do this, the GAM is constructed similarly as in Application 1 but with the
smoothing term specified as `te(latitude, longitude, time, d=c(2,1))` where
`d=c(2,1)` indicates that `latitude` and `longitude` share the same dimensions
(i.e., both are spatial and in units of degrees) while the `time` data are in
separate units. Note that results presented in Application 2 and 3 are based on
county-level data from the contiguous US excluding Alaska and Hawaii.

Using GAMs to present spatiotemporally smoothed estimates of COVID-19 mortality
allows for the synthesis of trends over time. Whereas the raw rates of COVID-19
mortality are noisy due to 72% of the land area in the US being classified as
rural and low population density [@health_resources__services_administration_defining_2022], the spatiotemporally smoothed GAM estimates
use population weighting via the offset specified to estimate a latent surface
that represents localized mortality rate averages in spatial coordinates and in
time.

Figure \@ref(fig:crude-vs-smoothed) shows the difference between crude mortality
rates and spatially smoothed mortality rates in January 2022 to illustrate the
level of noise present in raw rates and how spatially smoothing synthesizes
local geographic patterns into interpretable trends that can be meaningfully
interpreted as local area mortality risk levels pooling information across
nearby county rates.

Figure \@ref(fig:spatiotemporal-panel-figure) shows the results from a
spatiotemporally smoothed model in select months to highlight how spatiotemporal
smoothing can yield results that intuitively synthesize trends in space and
time.

@@

```{r spatiotemporal-panel-figure, fig.cap = "Spatiotemporally smoothed COVID-19 mortality estimates from a GAM fit to data from March 2020-August 2022. Panels show months selected to highlight the changing spatial patterns of COVID-19 mortality risk over time. An animated version is available online at \\url{https://github.com/ctesta01/covid.gradient.estimation/blob/main/analysis/09_spatiotemporal_models/animation/readme.md}"}
knitr::include_graphics(here("analysis/09_spatiotemporal_models/temporal_panel_figure.png"))
```

\newpage
## Application 3: Estimating Covariate Effects Adjusted for Spatiotemporal Autocorrelation

The final application of GAMs presented here is to estimate the effects of
covariates after adjusting for a spatiotemporally smoothed term as shown in
Application 2. This has the interpretation of asking what are the changes to
COVID-19 mortality rates associated with each covariate after taking into
account local geographic patterns over time. In particular, this is relevant for
understanding what county-level measures are associated with elevated COVID-19
mortality even after adjusting for where COVID-19 rates were locally elevated
or depressed due to variation in the timing of local introduction, transmission,
and diffusion.

```{r neighboring-graph-morans-i, fig.cap = "The spatial autocorrelation of COVID-19 rates in January 2022. Panel A shows the graph of which US counties are connected by neighboring relations. Panel B shows a scatterplot depicting each US counties' COVID-19 rate (x-axis) compared to the average of its neighbors' COVID-19 mortality rates (y-axis). A regression line shows the association between counties' COVID-19 rates and their neighboring counties' COVID-19 rates.  Dotted lines indicated the average for US county-level crude COVID-19 mortality rates (vertical) and for the average of each counties' neighboring counties' COVID-19 rates (horizontal)."}
knitr::include_graphics(here("analysis/08_spatiotemporal_autocorrelation/county_network_and_morans_plot.png"))
```

```{r temporal-acf, fig.cap = "The median (black) and quantiles (shaded bands) of temporal autocorrelation of counties' COVID-19 mortality rates over time starting from April 2020.", out.width = '50%', fig.align='center'}
knitr::include_graphics(here("analysis/08_spatiotemporal_autocorrelation/temporal_acf.png"))
```

To motivate the need for and application of spatiotemporal smoothing in
assessing the associated changes in COVID-19 mortality with county-level
covariates, Figure \@ref(fig:neighboring-graph-morans-i) presents the Moran's I
diagnostic plot for January 2022 which summarizes the amount of autocorrelation
between counties' COVID-19 mortality rates and a population-weighted average of
each counties' neighboring counties' COVID-19 mortality rates. The observed
autocorrelation indicates that there is an association between counties'
COVID-19 rates and their neighboring counties' COVID-19 rates, implying that
without taking this patterning into account, regression models that do not
explicitly model the effect of spatial autocorrelation may be biased due to
inappropriately assuming that county-level data are independent of one another. Second, 
while the Moran's I plot demonstrates the spatial-correlation present in the data
at a single time-point, the following Figure \@ref(fig:temporal-acf) shows the
tendency of county COVID-19 observations to be autocorrelated between subsequent
months and moreover there appears to be an oscillating trend at the 6-month and
12-month marks suggestive of effects due to the dynamics associated with
temporary infection induced immunity and seasonality.

```{r with-and-without-spatiotemporal-adjustment, fig.cap = "Estimates of increases in COVID-19 mortality associated with county-level covariates without (first) and with (second) adjustment for spatiotemporal autocorrelation. Panels A and B show COVID-19 mortality increases associated with varying levels of median age over time; C and D show associations with log population density; E and F median income; G and H for proportion in poverty; I and J the ICE for racialized economic segregation; K and L political lean.", out.width = '85%', fig.align='center'}
knitr::include_graphics(here("analysis/09_spatiotemporal_models/panel_figure.png"))
```

Figure \@ref(fig:with-and-without-spatiotemporal-adjustment) shows the effects associated
with each covariate over time with and without adjustment for spatiotemporal trends. 
The models for population density and median income show particularly marked changes
with the spike at high population density and high median incomes disappearing 
after adjusting for a spatiotemporally smoothed term. This likely reflects the
early emergence and surge of COVID-19 in the greater New York City area as being
predominantly a feature of the combined effect of human geography and infectious
disease dynamics where the earliest introductions of COVID-19 were made to dense
social networks in the US. Other effects remain qualitatively changed very little,
suggesting that the trends in COVID-19 mortality associated with these county-level
measures remains durable when taking into account spatiotemporal autocorrelation.

\newpage
# Discussion

This paper presents three applications of GAMs to county-level COVID-19
mortality data from the US during March 2020 to August 2022. This paper
addresses an important need, which is to assess what measures of the
socioeconomic and geopolitical landscape are associated with increases in
COVID-19 mortality in a way that takes into account how infectious disease
dynamics are locally correlated in space and time. However, the approaches
outlined in this paper are not without several limitations which include (at
least) the following: the potential for cross-level confounding, lack of
individual level data, an inability to include all possible metrics, further
improvements to the methods used may be warranted, and we are unable to make
claims about the causal relationships between the variables investigated.

In analyses that use area-level aggregated data, the ecologic fallacy refers to
the possibility that associations between area-level measures and outcomes may
not reflect the associations that would be observed if analyzing individual
level data on the same measures and outcomes as a result of cross-level
confounding [@greenland_ecologic_2001]. It's important to note that
some of the variables are included to contextualize the places in which people
live, reside, and work and area-based social metrics should be included in
individual-level analyses so as to accurately capture the effects of both
individual-level risk factors and risk factors that stem from the contexts in
which people exist [@testa_public_2022].

One particular improvement that could be made in future work is to address age
adjustment by using direct or indirect age standardization [@anderson_age_1998] or 
to create age-group stratified models. In the applications presented here,
county-level median age is adjusted for in order to control for age effects.

Additional variables like vaccination rates, mobility data, further variables
that relate to risk factors for COVID-19, and other COVID-19 related outcomes
like case trends and hospitalizations may be worthwhile to investigate under a
similar framework. The emphasis on trends in COVID-19 mortality in the
applications in this paper was done to prioritize presenting data that are less
subject to the reporting inconsistencies in COVID-19 cases and
hospitalizations [@galaitsi_challenges_2021].

Further improvements to the methods employed here could include use of soap film
smoothing splines which are more appropriate in settings where non-convex
geographic boundaries are common (as in with peninsulas and bays)
[@wood_soap_2008]. Taking a critical eye to the assumptions employed reveals yet
another area of potential improvement, which is the nature of how the complexity of the 
smoothing splines employed are penalized in the GAM framework. It is conceivable that
geographic changes in COVID-19 mortality rates are not equally smooth across 
all regions of the US, potentially warranting rapid changes in some areas 
and smooth changes in other areas. The Bayesian wombling literature provides
the means to identify areas of rapid change in latent surface models and may pose
a fruitful direction of exploration for future efforts to understand trends
in COVID-19 mortality rates in the US [@gelfand_bayesian_2015].

@@

This work introduces generalized additive models (GAMs) applied to COVID-19
mortality data to document the disparities and inequities in which US counties
at what times suffered the worst mortality outcomes. The effort to document
inequities and explain the driving mechanisms behind them is crucial in building
up the evidence necessary to enact policy changes that can mitigate unjust and
preventable harm in the future. To this end, it is critical that efforts to
model infectious disease dynamics take into account how disease transmission and
outcomes are spatiotemporally patterned, violating any misplaced assumptions
that area-level disease rates can be treated as independent observations. By
presenting novel applications of GAMs to trends in COVID-19 mortality, this
paper illustrates new ways of understanding and visualizing the disproportionate
burden some communities have suffered incorporating complex patterns of
interaction across socioeconomic and geopolitical variables and spatiotemporal
trends.

\newpage
# References
